---
title: "Frequently asked questions"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FAQ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{css echo=FALSE}
img {
    border: 0px !important;
    margin: 2em 2em 2em 2em !important;
}
code {
    border: 0px !important;
}
```

```{r echo=FALSE, results="hide"}
knitr::opts_chunk$set(
    cache = FALSE,
    echo = TRUE,
    collapse = TRUE,
    comment = "#>"
)
options(clustermq.scheduler = "local")
suppressPackageStartupMessages(library(clustermq))
```

## Installation errors {#install}

To compile this package a fully C++11 compliant compiler is required. This is
[implicit for CRAN packages](https://www.tidyverse.org/blog/2023/03/cran-checks-compiled-code/)
since `R=3.6.2` and is hence not listed in _SystemRequirements_. If you
encounter an error saying that that no matching function call to
`zmq::message_t::message_t(std::string&)` exists, your compiler does not
(fully) support this and the automated check failed for some reason.

```{sh eval=FALSE}
In file included from CMQMaster.cpp:2:0:
CMQMaster.h: In member function ‘void CMQMaster::proxy_submit_cmd(SEXP, int)’:
CMQMaster.h:146:40: error: no matching function for call to ‘zmq::message_t::message_t(std::string&)’
         mp.push_back(zmq::message_t(cur));
```

This happens for instance for old versions of the `gcc` compiler (default on
most Linux distributions). You can check your version in the terminal using:

```{sh eval=FALSE}
# the minimum required gcc version is 5.5 for full C++11 support (3.3 for clang)
cc --version
```

In this case, it is _very_ likely that your HPC system already has a newer
compiler installed that you need to add to your `$PATH` or load as a module.
Once this is set, you can install the package from R *that was started in a
terminal that has this module/path active*.

## Session gets stuck at "Running calculations" {#stuck}

Your R session may be stuck at something like the following:

```{r eval=FALSE}
> clustermq::Q(identity, x=42, n_jobs=1)
Submitting 1 worker jobs (ID: cmq8480) ...
Running 1 calculations (5 objs/19.4 Kb common; 1 calls/chunk) ...
```

You will see this every time your jobs are queued but not yet started.
Depending on how busy your HPC is, this may take a long time. You can check the
queueing status of your jobs in the terminal with _e.g._ `qstat` (SGE), `bjobs`
(LSF), or `sinfo` (SLURM).

If your jobs are already finished, this likely means that the `clustermq`
workers can not connect to the main session. You can confirm this by passing
[`log_worker=TRUE`](https://mschubert.github.io/clustermq/articles/userguide.html#debugging-workers)
to `Q` and inspect the logs created in your current working directory. If they
state something like:

```{sh eval=FALSE}
> clustermq:::worker("tcp://my.headnode:9091")
2023-12-11 10:22:58.485529 | Master: tcp://my.headnode:9091
2023-12-11 10:22:58.488892 | connecting to: tcp://my.headnode:9091:
Error: Connection failed after 10016 ms
Execution halted
```

the submitted job is indeed unable to establish a network connection with the
head node. This can happen if your HPC does not allow incoming connections, but
more likely happens because there are multiple network interfaces, only some of
which have access to the head node.

You can list the available network interfaces using the `ifconfig` command in
the terminal. Find the interface that shares a subnetwork with the head node
and add the [R option](https://mschubert.github.io/clustermq/articles/userguide.html#options) `clustermq.host=<interface>`. If this is
unclear, contact your system administrators to see which interface to use.

## SSH not working {#ssh}

Before trying remote schedulers via SSH, make sure that the scheduler works
when you first connect to the cluster and run a job from there.

If the terminal is stuck at

```
Connecting <user@host> via SSH ...
```

make sure that each step of your SSH connection works by typing the following
commands in your **local** terminal and make sure that you don't get errors or
warnings in each step:

```{sh eval=FALSE}
# test your ssh login that you set up in ~/.ssh/config
# if this fails you have not set up SSH correctly
ssh <user@host>

# test port forwarding from 54709 remote to 6687 local (ports are random)
# if the fails you will not be able to use clustermq via SSH
ssh -R 54709:localhost:6687 <user@host> R --vanilla
```

If you get an `Command not found: R` error, make sure your `$PATH` is set up
correctly in your `~/.bash_profile` and/or your `~/.bashrc` (depending on your
cluster config you might need either). You may also need to modify your [SSH
template](https://mschubert.github.io/clustermq/articles/userguide.html#ssh-template)
to load R as a module or conda environment.

If you get a SSH warning or error try again with `ssh -v` to enable verbose
output. If the forward itself works, run the following in your local R session
(ideally also in command-line R, [not only in
RStudio](https://github.com/mschubert/clustermq/issues/206)):

```{r eval=FALSE}
options(clustermq.scheduler = "ssh",
        clustermq.ssh.log = "~/ssh_proxy.log")
Q(identity, x=1, n_jobs=1)
```

This will create a log file *on the remote server* that will contain any errors
that might have occurred during `ssh_proxy` startup.

If the `ssh_proxy` startup fails on your local machine with the error

```
Remote R process did not respond after 5 seconds. Check your SSH server log.
```

but the server log does not show any errors, then you can try increasing the
timeout:

```{r eval=FALSE}
options(clustermq.ssh.timeout = 30) # in seconds
```

This can happen when your SSH startup template includes additional steps before
starting R, such as activating a module or conda environment, or having to
confirm the connection via two-factor authentication.

## Running the master inside containers {#master-in-container}

If your master process is inside a container, accessing the HPC scheduler is
more difficult. Containers, including singularity and docker, isolate the
processes inside the container from the host. The *R* process will not be able
to submit a job because the scheduler cannot be found.

Note that the HPC node running the master process must be allowed to submit
jobs. Not all HPC systems allow compute nodes to submit jobs. If that is the
case, you may need to run the master process on the login node, and discuss the
issue with your system administrator.

If your container is binary compatible with the host, you may be able to bind
in the scheduler executable to the container.

For example, PBS might look something like:

```{sh eval=FALSE}
#PBS directives ...

module load singularity

SINGULARITYENV_APPEND_PATH=/opt/pbs/bin
singularity exec --bind /opt/pbs/bin r_image.sif Rscript master_script.R
```

A working example of binding SLURM into a CentOS 7 container image from a
CentOS 7 host is available at
https://groups.google.com/a/lbl.gov/d/msg/singularity/syLcsIWWzdo/NZvF2Ud2AAAJ

Alternatively, you can create a script that uses SSH to execute the scheduler
on the login node. For this, you will need an SSH client in the container,
[keys set up for password-less login](https://www.digitalocean.com/community/tutorials/how-to-configure-ssh-key-based-authentication-on-a-linux-server),
and create a script to call the scheduler on the login node via ssh (e.g.
`~/bin/qsub` for SGE/PBS/Torque, `bsub` for LSF and `sbatch` for Slurm):

```{sh eval=FALSE}
#!/bin/bash
ssh -i ~/.ssh/<your key file> ${PBS_O_HOST:-"no_host_not_in_a_pbs_job"} qsub "$@"
```

Make sure the script is executable, and bind/copy it into the container
somewhere on `$PATH`. Home directories are bound in by default in singularity.

```{sh eval=FALSE}
chmod u+x ~/bin/qsub
SINGULARITYENV_APPEND_PATH=~/bin
```
